CrypTool Script \\ Mathematics and Cryptography

### 2003-03-27 script/en/chapters/numbertheory.tex
be updates
more be updates

Examples of the decomposition of such numbers into prime factors:
\begin{eqnarray*}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1,029 & = & 3 * 7^3  \nonumber \\
5,324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray*}

@@

For example, $60 = 2*2*3*5 = 2^2*3*5$. And this --- other than changing the order of the factors --- is the only 
way in which the number $60$ can be factorised.

@@

If $J(m)$ cannot be determined\footnote{According to the first fundamental theorem of number 
theory\index{Number theory!fundamental theorem} and theorem \ref{thm-zth-phinum}, we determine $J(m)$ by reducing $m$ to prime factors.}, it is difficult to calculate 
the $c$-th root. This forms the basis for the security assumption used by the RSA encryption system (see 
Sub-section 4.3.1: \hyperlink{RSAVerfahren}{the RSA procedure} and \hyperlink{RSABeweis}{Sub-section 3.10 Proof of the RSA procedure}).
\end{itemize}
The time required for inverting addition and multiplication, on the other hand, 
is simply proportional to $\log m$ or $(\log m)^2$. Powers (for a number $x$ 
calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions 
(See Overview of the one way functions in this \hyperlink{OneWayFunktion1}{Script} and
\hyperlink{OneWayFunktion2}{artikel}). \index{One way function}

@@

\begin{definition}\label{def-zth-phiofn} \hypertarget{EulerFunction}{} 
The Euler function\footnote{Is often described as the Euler phi function\index{Euler!(phi) function} $\Phi(n)$.} $J(n)$ specifies the number of 
elements in $\mathbb{Z}_n^*$.
\end{definition}
$J(n)$ also specifies how many whole numbers have multiplicative inverses in mod $n$.
$J(n)$ can be calculated if we know the prime factors of $n$.

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler}\index{RSA}
In order to prove the RSA procedure, we need Fermat's theorem and its generalisation (Euler-Fermat theorem).

\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, 17.8.1601 -- 12.1.1665.\index{Fermat}}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA procedure} is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
Within literature and movies not only classical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{itemize}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
  \index{Elsner 1999}, 1999
  (included in the CrypTool package\index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\end{itemize}
} 
procedure in the group $\mathbb{Z}_n^*$.

@@

In order to ensure that this system works for every message $M$, the following four requirements must be met:
\begin{itemize}
\item[{\bf 1.}] $D ( S;  E (P; M) ) = M$  for every $M$  (invertibility) and $M$ takes ``very many'' of its possible values.
\item[{\bf 2.}] All $(S, P)$ pairs are different for all participants (i.e. lots of them are needed).
\item[{\bf 3.}] The time required to derive $S$ from $P$ is at least as high as the time required to decrypt $M$ with no knowledge of $S$.
\item[{\bf 4.}] Both $C$ and $M$ can be calculated relatively easily.
\end{itemize}
The 1st requirement is a general condition for all cryptographic encryption algorithms.

@@

restriction is fulfilled (see
\href{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}
{\tt http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}).

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{``Primes in P'': Primality testing is polynomial} \label{TWIRLDevice} \index{TWIRL Device}\mbox{}

@@

The web page
\href{http://www.crypto-world.com}{\texttt{http://www.crypto-world.com}}\ 
contains an excellent overview about the factoring records of composed
integers using different methods.  The current record (as of June 2002)
obtained using the GNFS method (General Number Field Sieve) factorized a
158-digit into its both prime factors (these are build with 73 and 86
decimal digits).

This record, established on January 18, 2002 by researchers of the German
university of Bonn (see the press statement (in German language): \\
\href{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}
{\texttt{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}}), got much less attention within the press than the
solution of the RSA challenge, established on August 22, 1999, when
researchers from the Netherlands factorized the 155-digit number into its
both 78-digit primes (see Chapter \ref{chptSecurityParam}). 
One reason is that 512 bit RSA-155 was a kind of {\em magic} border.

@@

The 6 smaller prime factors, already found before have been:
$$
\begin{array}{c}
3, 1907, 425796183929, \\
1624700279478894385598779655842584377, \\
3802306738549441324432139091271828121 \quad {\rm and} \\
128064886830166671444802576129115872060027.
\end{array}
$$
The first 3 factors can be easily computed \footnote{%
e.g. using CrypTool\index{CrypTool} via menu 
{\bf Indiv. Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.
}
The next three prime factors were found by P.~Zimmerman
(\href{http://www.loria.fr/~zimmerma/ecmnet}{\tt http://www.loria.fr/\~{}zimmerma/ecmnet}), 
T.~Grandlund (\href{http://www.swox.se/gmp/}{\tt http://www.swox.se/gmp/})
and R. Harley during the years 1999 and 2000 using the elliptic curve factorisation method.

@@

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling} K. Rowling\footnotemark\:}\\
It is our choices, that show what we truly are, far more than our abilities.}}
\end{center}

@@

Whitfield Diffie, Martin E. Hellman and Ralph Merkle developed this DH key 
exchange protocol in Stanford in 1976\footnote{%
With CrypTool\index{CrypTool} v1.3.04  this exchange protocol has been
visualized: you can execute the single steps with concrete numbers using 
menu {\bf Indiv. Procedures \textbackslash{} Diffie-Hellman Demonstration}.
}
.

@@

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice):
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.}
\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.
\vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +6 pt
} is extremely difficult to solve.

@@

Before applying the RSA procedure to a text, we will first demonstrate it
directly using a single number as message
\footnote{Using CrypTool\index{CrypTool} v1.3 you can solve this with
the {\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} RSA Cryptosystem}.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should lie between $11$ and $40$ and must be relatively prime to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1 {\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

\bibitem[GISA2002]{GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency) \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \href{http://www.bsi.de/aufgaben/projekte/pbdigsig/download/kryptalg.pdf}
    {\tt http://www.bsi.de/aufgaben/projekte/pbdigsig/download/kryptalg.pdf} \\
    A statement on these recommendations: \\
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}

@@

\bibitem[Schneier1996]{Schneier1996} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley, 2nd edition 1996.

@@

   \item Factorisation records \index{Factorisation records}\\
         \href{http://www.crypto-world.com/}
         {\texttt{http://www.crypto-world.com/}} \\
         \href{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}
         {\texttt{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}}

@@

I would like to take this opportunity to thank 
\begin{itemize}
  \item {Mr. Henrik Koy for making many very useful suggestions, 
         for very constructive proof-reading this article and
         for helping with TeX.}
  \item {J\"org-Cornelius Schneider for his TeX support.}
  \item {Dr. Illies for pointing me to Pari-GP \index{Pari-GP}.}
\end{itemize}

>>>>

Examples of the decomposition of such numbers into prime 
factors\index{Prime factor!decomposition}:
\begin{eqnarray*}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1,029 & = & 3 * 7^3  \nonumber \\
5,324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray*}

@@

For example, $60 = 2*2*3*5 = 2^2*3*5$. And this --- other than changing the order of the factors --- is the only 
way in which the number $60$ can be factorised\index{Prime factor}.

@@

If $J(m)$ cannot be determined\footnote{According to the first fundamental theorem of number 
theory\index{Number theory!fundamental theorem} and theorem \ref{thm-zth-phinum}, we determine $J(m)$ by reducing $m$ to prime factors\index{Prime factor!decomposition}.}, it is difficult to calculate 
the $c$-th root. This forms the basis for the security assumption used by the RSA encryption system (see 
Sub-section 4.3.1: \hyperlink{RSAVerfahren}{the RSA procedure} and \hyperlink{RSABeweis}{Sub-section 3.10 Proof of the RSA procedure}).
\end{itemize}
The time required for inverting addition and multiplication, on the other hand, 
is simply proportional to $\log m$ or $(\log m)^2$. Powers (for a number $x$ 
calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions 
(See Overview of the one way functions in this \hyperlink{OneWayFunktion1}{Script} and
\hyperlink{OneWayFunktion2}{artikel}). \index{One way function}

@@

\begin{definition}\label{def-zth-phiofn} \hypertarget{EulerFunction}{} 
The Euler function\footnote{Is often described as the Euler phi function\index{Euler!(phi) function} $\Phi(n)$.} $J(n)$ specifies the number of 
elements in $\mathbb{Z}_n^*$.
\end{definition}
$J(n)$ also specifies how many whole numbers have multiplicative inverses in mod $n$.
$J(n)$ can be calculated if we know the prime factors of $n$\index{Prime factor!decomposition}.

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler}\index{RSA}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please also see chapter 2.5}.

\hypertarget{KleinerSatzFermat-chap3}{}
\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, 17.8.1601 -- 12.1.1665.\index{Fermat}}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA procedure} is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
Within literature and movies not only classical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{list}{\textbullet}{\leftmargin10pt}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner\index{Elsner 1999}, 1999
  (included in the CrypTool package\index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\end{list}\vspace{-\baselineskip}\vspace{-\parskip}}
procedure in the group $\mathbb{Z}_n^*$.

@@

In order to ensure that this system works for every message $M$, the following four {\bf requirements} must be met:
\begin{itemize}
\item[{\bf 1.}] $D ( S;  E (P; M) ) = M$  for every $M$  (invertibility) and $M$ takes ``very many'' of its possible values.
\item[{\bf 2.}] All $(S, P)$ pairs are different for all participants (i.e. lots of them are needed).
\item[{\bf 3.}] The time required to derive $S$ from $P$ is at least as high as the time required to decrypt $M$ with no knowledge of $S$.
\item[{\bf 4.}] Both $C$ and $M$ can be calculated relatively easily.
\end{itemize}
The 1st requirement is a general condition for all cryptographic encryption algorithms.

@@

restriction is fulfilled (see \cite{GISA2002}).

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{``Primes in P'': Primality testing is polynomial} \label{PrimesinP} \index{Primality testing}\mbox{}

@@

The web page
\href{http://www.crypto-world.com}{\texttt{http://www.crypto-world.com}}\ 
contains an excellent overview about the factoring 
records\index{factoring records} of composed
integers using different methods.
\hypertarget{C158-chap3}{}
The current record (as of June 2002) obtained using the GNFS method
(General Number Field Sieve) factorized a 158-digit into its both prime
factors (these are build with 73 and 86 decimal digits).

This record, established on January 18, 2002 by researchers of the German
university of Bonn (see: \\
\href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
{\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}}), got much less attention within the press than the
solution of the RSA challenge, established on August 22, 1999, when
researchers from the Netherlands factorized the 155-digit number into its
both 78-digit primes (see Chapter \ref{chptSecurityParam}). 
One reason is that 512 bit RSA-155 was a kind of {\em magic} border.

@@

The 6 smaller prime factors, already found before have been:
$$
\begin{array}{c}
3, 1907, 425796183929, \\
1624700279478894385598779655842584377, \\
3802306738549441324432139091271828121 \quad {\rm and} \\
128064886830166671444802576129115872060027.
\end{array}
$$
The first 3 factors can be easily computed\footnote{%
e.g. using CrypTool\index{CrypTool} via menu 
{\bf Indiv. Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.
}.
The next three prime factors were found by P.~Zimmerman
(\href{http://www.loria.fr/~zimmerma/ecmnet}{\tt http://www.loria.fr/\~{}zimmerma/ecmnet}), 
T.~Grandlund (\href{http://www.swox.se/gmp/}{\tt http://www.swox.se/gmp/})
and R. Harley during the years 1999 and 2000 using the elliptic curve factorisation method.

@@

As you notice the factorized compound numbers built of 2 prime factors are
much smaller than the especially structured numbers, for which primality 
tests\index{Primality testing}} are able to decide whether they are prime 
or not (see chapter 2.4 and 2.5).

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling} K. Rowling\footnotemark:}\\
It is our choices, that show what we truly are, far more than our abilities.}}
\end{center}

@@

Whitfield Diffie, Martin E. Hellman and Ralph Merkle developed this DH key 
exchange protocol in Stanford in 1976\footnote{%
With CrypTool\index{CrypTool} v1.3.04  this exchange protocol has been
visualized: you can execute the single steps with concrete numbers using 
menu {\bf Indiv. Procedures \textbackslash{} Diffie-Hellman Demonstration}.
}.

@@

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice):
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.
}${}^,$\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.
\vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +6 pt
} is extremely difficult to solve.

@@

Before applying the RSA procedure to a text, we will first demonstrate it
directly using a single number as message\footnote{Using CrypTool\index{CrypTool} v1.3 you can solve this with
the {\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} RSA Cryptosystem}.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should lie between $11$ and $40$ and must be relatively prime to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1 {\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

\bibitem[GISA2002]{GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency) \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}
    {\tt http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf} \\
    A statement on these recommendations: \\
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}

@@

\bibitem[Schneier1996]{Schneier1996} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley and Sons, 2nd edition 1996.

@@

   \item Factorisation records and challenges \index{Factorisation records}\\
         \href{http://www.crypto-world.com/}
         {\texttt{http://www.crypto-world.com/}} \\
         \href{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}
         {\texttt{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}} \\
         \href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
         {\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}} \\
         \href{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}
         {\texttt{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}}

@@

I would like to take this opportunity to thank 
\begin{itemize}
  \item {Mr. Henrik Koy for making many very useful suggestions, 
         for very constructive proof-reading this article and
         for helping with TeX. \\
	 Mr. Koy developed the functions and the complex dialog box of the
	 RSA cryptosystem, which allows you to execute the RSA samples
	 of this article.}
  \item {J\"org-Cornelius Schneider for his TeX support und for the many 
         cases where he helped when facing programming or design problems.}
  \item {Dr. Illies for pointing me to Pari-GP \index{Pari-GP}.}
\end{itemize}
